---
title: "model_predictors_breeding_container_definition"
author: "K Tiley"
date: "2026-06-19"
output: html_document
---

```{r setup, include=FALSE}

library(dplyr)
library(sf)
library(mgcv)
library(terra)
library(exactextractr)

```

## Minimal wMel spatial setup for sensitivity analysis

#### Wolbachia raster cells and flight buffers

```{r wmel, warning=FALSE}

wmel_rio <- read.csv("data_wmel_rio.csv")

wmel_rio_releases <- wmel_rio %>%
  filter(t >= 92) %>%
  mutate(
    date = as.Date(month),
    year = as.integer(format(date, "%Y")),
    month = as.integer(format(date, "%m"))
  )

wmel_points <- st_as_sf(
  wmel_rio_releases,
  coords = c("x", "y"),
  crs = 32723
)

make_square <- function(pt, size) {
  xy <- st_coordinates(pt)
  h <- size / 2
  
  st_polygon(list(matrix(c(
    xy[1] - h, xy[2] - h,
    xy[1] + h, xy[2] - h,
    xy[1] + h, xy[2] + h,
    xy[1] - h, xy[2] + h,
    xy[1] - h, xy[2] - h
  ), ncol = 2, byrow = TRUE)))
}

make_square_sf <- function(sf_points, size) {
  st_sf(
    sf_points,
    geometry = st_sfc(
      lapply(st_geometry(sf_points), make_square, size = size),
      crs = st_crs(sf_points)
    )
  )
}

# 500 × 500 m raster cells
wmel_rio_releases_sf_xy_buffer <- make_square_sf(wmel_points, size = 500)

unique_wmel_cells <- wmel_rio_releases_sf_xy_buffer %>%
  select(cell.id) %>%
  distinct()

# 700 × 700 m buffers
flight_buffers <- make_square_sf(wmel_points, size = 700)

unique_flight_buffers <- flight_buffers %>%
  transmute(cell.id, geometry) %>%
  distinct()

# Remove central 500 m cell to create hollow neighbouring buffer
buffer_intersection_same_cell <- st_intersection(
  unique_flight_buffers %>% rename(cell.id_flight = cell.id),
  unique_wmel_cells
) %>%
  filter(cell.id_flight == cell.id)

hollow_flight_buffers <- unique_flight_buffers %>%
  rename(cell.id_flight = cell.id) %>%
  st_difference(buffer_intersection_same_cell) %>%
  filter(cell.id_flight == cell.id) %>%
  transmute(cell.id = cell.id_flight, geometry) %>%
  st_make_valid()

```

#### Neighbouring %wMel

```{r neighbouring_wmel, warning=FALSE}

cell_areas <- wmel_rio_releases_sf_xy_buffer %>%
  mutate(area_cell = as.numeric(st_area(geometry))) %>%
  st_drop_geometry() %>%
  distinct(cell.id, area_cell)

intersection_sf <- st_intersection(
  hollow_flight_buffers %>%
    transmute(cell.id_flight = cell.id, geometry),
  wmel_rio_releases_sf_xy_buffer %>%
    select(cell.id, year, month, wmel_pctg)
) %>%
  mutate(area_intersection = as.numeric(st_area(geometry))) %>%
  left_join(cell_areas, by = "cell.id") %>%
  mutate(
    weight = area_intersection / area_cell,
    weighted_wmel_pctg = wmel_pctg * weight
  )

average_neighbouring_wmel_values <- intersection_sf %>%
  st_drop_geometry() %>%
  group_by(cell.id_flight) %>%
  summarise(
    neighbouring_wmel_pctg_mean_avg =
      sum(weighted_wmel_pctg, na.rm = TRUE) /
      sum(weight, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  rename(cell.id = cell.id_flight)

```

#### Water tank density
Source: Steffen Knoblauch

```{r water_tank_density, warning=FALSE}

water_tank <- terra::rast("water_tank_density.tif")

water_tank_values <- unique_wmel_cells %>%
  st_transform(terra::crs(water_tank)) %>%
  select(cell.id) %>%
  mutate(
    water_tank_density = exactextractr::exact_extract(
      water_tank,
      .,
      fun = "mean"
    )
  ) %>%
  st_drop_geometry()

```

#### Breeding container counts
Source: Steffen Knoblauch

```{r breeding_container_counts, warning=FALSE}

# Import container point data once
breeding_container_types <- st_read("GSV_container_counts.json", quiet = TRUE) %>%
  st_transform(32723)

# Count each container type per raster cell
breeding_container_types_values <- unique_wmel_cells %>%
  st_transform(st_crs(breeding_container_types)) %>%
  select(cell.id) %>%
  st_join(breeding_container_types) %>%
  st_drop_geometry() %>%
  group_by(cell.id) %>%
  summarise(
    across(where(is.numeric), ~ sum(.x, na.rm = TRUE)),
    .groups = "drop"
  )

```


## Test water tank counts vs density

```{r water_tank_count_vs_density, warning=FALSE}

# Water tank count
m_wt_count_rhs <- ~ s(neighbouring_wmel_pctg_mean_avg) + water_tank

m_wt_count <- gam(
  make_multinom_flist(m_wt_count_rhs),
  family = multinom(K = K),
  method = "REML",
  data = m_df
)

# Water tank density
m_wt_density_rhs <- ~ s(neighbouring_wmel_pctg_mean_avg) + water_tank_density

m_wt_density <- gam(
  make_multinom_flist(m_wt_density_rhs),
  family = multinom(K = K),
  method = "REML",
  data = m_df
)

AIC(m_wt_count, m_wt_density)

```

Conclusion:
Water tank density showed stronger support than water tank count and was retained for subsequent comparisons.


## Test individual container types

```{r test_container_type, warning=FALSE}

# Combine candidate carrying-capacity predictors
carrying_capacity <- average_neighbouring_wmel_values %>%
  left_join(breeding_container_types_values, by = "cell.id") %>%
  left_join(water_tank_values, by = "cell.id")

carrying_capacity_predictors <- names(carrying_capacity %>% select(-cell.id))

carrying_capacity_predictors_no_spatial <- setdiff(
  carrying_capacity_predictors,
  c("neighbouring_wmel_pctg_mean_avg", "water_tank")
)

# Import outcome
k_wmel <- read.csv("k4_reassigned_wmel_all_2025-09-24.csv", header = TRUE) %>%
  select(cell.id, cluster) %>%
  filter(!is.na(cluster)) %>%
  distinct(cell.id, .keep_all = TRUE)

# Combine outcome and predictors
m_df <- left_join(k_wmel, carrying_capacity, by = "cell.id")

# Check this reference level exists before running
m_df$cluster <- as.factor(m_df$cluster)
m_df$cluster <- relevel(m_df$cluster, ref = "Unsuccessful")
m_df$cluster_code <- as.numeric(m_df$cluster) - 1L

K <- max(m_df$cluster_code)

base_rhs <- as.formula(
  paste(
    "~ s(neighbouring_wmel_pctg_mean_avg) +",
    paste(carrying_capacity_predictors_no_spatial, collapse = " + ")
  )
)

base_rhs_first <- update(base_rhs, paste("cluster_code ~ ."))
f_list <- list(base_rhs_first, base_rhs, base_rhs)

m_container_types <- gam(
  f_list,
  data = m_df,
  family = mgcv::multinom(K = K),
  method = "ML",
  select = TRUE,
  na.action = na.exclude
)

AIC(m_container_types)
summary(m_container_types)

```

Conclusion:
The individual container types with strongest signal were: potted_plant, storm_drain_with_grid, and water_tank.
These were carried forward into the main predictor-processing script as breeding_container_sum.

