---
title: "mulitnomial_gam_2026-05-15"
author: "K Tiley"
date: "2026-05-25"
output: html_document
---

```{r setup, include=FALSE}

library(dplyr)            # pipelines
library(tidyr)            # dataframe manipulation
library(reshape2)         # dataframe manipulation
library(forcats)          # fct_recode()
library(nnet)             # multinomial logistic regression
library(mgcv)             # GAMs
library(caret)            # regression
library(ggplot2)          # plotting
library(sf)               # spatial data
library(spatialsample)    # spatial cross-validation
library(spdep)            # spatial autocorrelation
library(car)              # multicollinearity
library(purrr)            # iteration (map_* etc.)
library(rlang)
library(tibble)
library(stringr)
library(ggbeeswarm)

```


## Prepare data

```{r data_prep}

# upload predictors
predictors_df <- read.csv("model_predictors_df_updatedflightbarriers2025-09-30.csv", header = TRUE)
#names(predictors_df)

# upload outcome trajectories
wmel_reassigned <- read.csv("k4_reassigned_wmel_all_2025-09-24.csv", header = TRUE)  # outcome with reassignment for decreasing slopes
# sensitivity analysis so reassign cluster trajectories with original cluster from reassigned_ids in wmel_pctg_trajectories.Rmd
wmel_not_reassigned <- wmel_reassigned %>%
  mutate(
    cluster = case_when(
      cell.id %in% c(13, 22, 40, 60, 61, 170, 268) ~ "Moderate Success",
      cell.id %in% c(70, 292) ~ "High Success",
      TRUE ~ cluster
    )
  )

# prepare model dataframes
wmel_clustered_num <- wmel_not_reassigned %>%
  filter(!is.na(cluster)) %>%                                                                               # remove rows where cluster is NA
  mutate(cluster = factor(cluster, levels = c("Unsuccessful","Low Success","Moderate Success","High Success")),
         cluster_num = dplyr::recode(as.character(cluster), "Unsuccessful" = 3L, "Low Success" = 4L, "Moderate Success" = 2L, "High Success" = 1L,
                                     .default = NA_integer_, .missing = NA_integer_),
         release_zone = factor(release_zone)) %>%
  distinct(cell.id, .keep_all = TRUE) %>%                                                                   # retain one row (and cluster label) per cell.id
  dplyr::select(cell.id, cluster, cluster_num, x, y, release_zone)

wmel_model_df <- wmel_clustered_num %>%
  dplyr::select(cell.id, cluster, release_zone) %>%
  left_join(predictors_df, by = "cell.id") %>%
  mutate(across(
    .cols = -c(cell.id, cluster, release_zone),
    .fns = ~ replace_na(., 0)
    ),
    release_zone5 = release_zone,
    release_zone = case_when(
      release_zone5 %in% c("RJ3.1", "RJ3.2", "RJ3.3") ~ "RJ3",
      release_zone5 == "RJ2" ~ "RJ2",
      TRUE ~ "RJ1"
    ),
    release_zone = factor(release_zone)
  )

# predictor pool (exclude id, response, coords)
predictors_og <- setdiff(
  names(wmel_model_df),
  c("cell.id", "cluster", "neighbouring_wmel_pctg_mean_avg", "release_zone", "release_zone5")
)

```


## Check baseline spatial autocorrelation

Monte Carlo join-count test (for categorical variables, Moran's I is for continuous variables)
    Significant positive joins = same-cluster neighbours greater than expected under randomness = spatial clustering of the outcome

Note: using points instead of polygons is acceptable in this case as I set neighbour distance at 800m. For 500 x 500 m raster cells the distance to rook neighbours = 500m and queen neighbours = ~707m

```{r spatial_autocorrelation, warning=FALSE}

wmel_clustered_sf <- st_as_sf(wmel_clustered_num, coords = c("x", "y"), crs = 32723)                      # add geometries
#plot(st_geometry(wmel_clustered_sf), col = 'grey70')                                                      # sanity check:- geometries look ok
coords <- st_coordinates(wmel_clustered_sf)                                                               # get coords for spatial weighting
nb <- dnearneigh(coords, d1 = 0, d2 = 800)                                                                # distance-based neighbours: within 800 m
nb_k1 <- knn2nb(knearneigh(coords, k = 1))                                                                # list of neighbours number
nb_fix <- union.nb(nb, nb_k1)                                                                            # add one nearest-neighbour to any 'islands' (isolated cells)
cat("sanity check:- no zeros:\n")
table(card(nb_fix))
lw <- nb2listw(nb_fix, style = "W")                                                                       # build spatial weights
set.seed(1)
pvals <- c(                                                                                                # extracted p-value for TRUE (within specified cluster)
  joincount.mc(factor(wmel_clustered_sf$cluster %in% "Unsuccessful"), lw, nsim=999)[[2]]$p.value,
  joincount.mc(factor(wmel_clustered_sf$cluster %in% "Low Success"), lw, nsim=999)[[2]]$p.value,
  joincount.mc(factor(wmel_clustered_sf$cluster %in% "Moderate Success"), lw, nsim=999)[[2]]$p.value,
  joincount.mc(factor(wmel_clustered_sf$cluster %in% "High Success"), lw, nsim=999)[[2]]$p.value)
pvals_FDR <- p.adjust(pvals, "fdr")                                                                       # adjusted for False Discovery Rate (since doing multiple tests)
cat("\nFDR-adjusted join-count test p-values:\n")
print(pvals_FDR)

```
Statistical significance indicates strong spatial autocorrelation and justifies a smoothed spatial term.


## Predictor selection

#### a) Check for zero variance

```{r pred_variance}

nzv_vars <- nearZeroVar(wmel_model_df[, predictors_og], saveMetrics = TRUE) %>% filter(nzv == TRUE)
print(nzv_vars)

```

```{r pred_variance_refine}

# REMOVE zero variance predictors
nzv_vars_remove <- nzv_vars %>%
  filter(nzv == TRUE) %>%
  rownames()
predictors_clean1 <- setdiff(predictors_og, nzv_vars_remove)

```

#### b) Check for multicollinearity

```{r pred_multicollinearity}

cor_matrix <- cor(wmel_model_df[, predictors_clean1], use = "pairwise.complete.obs")
cor_matrix[upper.tri(cor_matrix, diag = TRUE)] <- NA                                                      # exclude diagonal and duplicates
cor_long <- as.data.frame(as.table(cor_matrix)) %>%                                                       # convert to long for easier filtering
  filter(!is.na(Freq)) %>%
  arrange(desc(abs(Freq)))
high_corr_pairs <- cor_long %>% filter(abs(Freq) > 0.8)                                                   # show pairs with |correlation| > 0.8
high_corr_pairs

```

```{r pred_multicollinearity_refine}

# REMOVE collinear predictors based on keeping most biologically important per pair
drops <- c("average_literacy","humidity_mean_idw","built_volume_density")
predictors_clean2 <- setdiff(predictors_clean1, drops)

```

```{r pred_multicollinearity_plot, fig.height=8, fig.width=10, warning=FALSE}

cor_matrix_drops <- cor(wmel_model_df[, predictors_clean2], use = "pairwise.complete.obs")
cor_matrix_drops[upper.tri(cor_matrix_drops, diag = TRUE)] <- NA                                          # exclude diagonal and duplicates
cor_melt <- melt(cor_matrix_drops, varnames = c("Var1", "Var2"))                                           # melt to long format
cor_melt <- cor_melt %>%
  mutate(label = ifelse(abs(value) >= 0.3, round(value, 2), ""))                                            # create column for labels where absolute value >= 0.6
corr_plot <- ggplot(cor_melt, aes(Var1, Var2, fill = value)) +                                              # plot corrplot with labelled correlation coefficients
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", 
                       midpoint = 0, limit = c(-1,1), space = "Lab") +
  geom_text(aes(label = label), color = "black", size = 3) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  coord_fixed() +
  ggtitle("Correlation Matrix (Labels > |0.3|)")
corr_plot

```

#### c) Check for high Variance Inflation Factors (VIF)

As a rule of thumb:
-  <5 low, 5-10 moderate
-  >10 high multicollinearity & consider removing or combining
-  >30 severe multicollinearity & must address as model is not valid

```{r pred_vif}

lm_vif_test <- lm(as.numeric(cluster) ~ ., data = wmel_model_df[, c("cluster", predictors_clean2)])
vif_values <- car::vif(lm_vif_test)
vif_values

```

Problem: proportion_residential & proportion_non_residential
Resolution: 'residualise' proportion_non_residential so that the non-residential buffer effect is retained without the non-residential data trends explained by proportion_residential
How: fit a model to get residuals, but which model? -> lm preferred unless curvature then gam needed

```{r tester}

# a) visually curves away from the line, indicates non-linear
ggplot(wmel_model_df, aes(x = proportion_residential, y = proportion_non_residential)) +
  geom_point(alpha = 0.25) +
  geom_smooth(method = "lm", se = FALSE, linetype = 2, linewidth = 0.8, color = "black") +
  geom_smooth(method = mgcv::gam, formula = y ~ s(x, k = 6), se = FALSE, linewidth = 1)
# b) the proportion_residential smooth term is significantly better than the linear predictor, indicates non-linear
m_lin <- gam(proportion_non_residential ~ proportion_residential, data = wmel_model_df, method = "ML")
m_spl <- gam(proportion_non_residential ~ s(proportion_residential, k = 10), data = wmel_model_df, method = "ML")
AIC(m_lin,m_spl)
# c) edf greater than 1 indicates non-linear, non-significant p-value indicates k isn't too small
gam.check(m_spl)


```

Evidence of non-linearity, so fit a GAM.

```{r pred_vif_refine}

# how: fit a quick model to get residuals -> gam
aux <- mgcv::gam(proportion_non_residential ~ s(proportion_residential, k = 6), data = wmel_model_df, method = "REML")
wmel_model_df$proportion_non_residential <- resid(aux)
# problem resolved in vif?
lm_vif_test <- lm(as.numeric(cluster) ~ ., data = wmel_model_df[, c("cluster", predictors_clean2)])
vif_values <- car::vif(lm_vif_test)
vif_values

```


## Scale predictors

Scale predictors for comparable coefficients which tell you the expected change in log-odds per 1 SD increase in the predictor. Method: subtract predictor mean and divide by SD.

```{r pred_scaled}

wmel_model_df_scaled <- wmel_model_df[, c(predictors_clean2, "cell.id", "cluster", "release_zone", "neighbouring_wmel_pctg_mean_avg")] %>%
  mutate(across(all_of(predictors_clean2), ~ as.numeric(scale(.x))))
#summary(wmel_model_df_scaled)

```


## Model screening

```{r screening_df}

# final model dataframe
wmel_model_df_scaled_subset <- wmel_model_df_scaled %>%
  dplyr::select(any_of(c(predictors_clean2, "cluster", "release_zone", "neighbouring_wmel_pctg_mean_avg")))
df <- droplevels(wmel_model_df_scaled_subset)

df$cluster <- relevel(df$cluster, ref = "Unsuccessful")
df$cluster_code <- as.numeric(df$cluster) - 1L                                                            # recode response to integers 0..K with desired baseline as 0
K <- max(df$cluster_code)
f_list0 <- list(cluster_code ~ s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone), ~ s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone), ~s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone))
m0 <- gam(f_list0, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)  # intercept-only reference model
aic_base <- AIC(m0)

```

```{r screening_helpers}

build_flist_linear <- function(
    rhs, K, resp = "cluster_code",
    k = 10, bs = "ts"
    ){

  # rhs  = RHS formula like ~ x1 + x2 + x3
  # K    = number of classes
  # resp = response variable name
  # k    = default basis size (scalar, default 10) OR a *named* numeric vec per var
  # bs   = basis for s(), default "ts" (thin-plate w/ shrinkage)
  
  stopifnot(inherits(rhs, "formula"), K >= 2)
  
  term_labels <- xfun::attr2(terms(rhs), "term.labels")
  
  pick_k <- function(v) {
    if (length(k) > 1 && !is.null(names(k)) && v %in% names(k)) k[[v]] else as.numeric(k)[1]
  }
  
  # always add the spatial proxy smooth
#  spatial_proxy <- "s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10)"
#  spatial_proxy <- "s(release_zone, bs='re')"
  spatial_proxy <- "s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone)"


  # assemble RHS (deduplicate, preserve first occurrence)
  rhs_terms <- unique(c(term_labels, spatial_proxy))
  rhs_new   <- as.formula(paste("~", paste(rhs_terms, collapse = " + ")))
  
  # mgcv multinomial formula list: first has response, rest are RHS
  c(list(update(rhs_new, paste(resp, "~ ."))),
    rep(list(rhs_new), K - 1))
}

```


## Forward selection by AIC |>2|

```{r forward_selection_1}

univariate_tbl_1 <- lapply(predictors_clean2, function(x){
  f1 <- as.formula(paste("~", x))
  list_f <- build_flist_linear(f1, K)
  m <- gam(list_f, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)
  data.frame(term = x, AIC  = AIC(m), dAIC = AIC(m) - aic_base)
}) %>% bind_rows() %>% arrange(AIC)
head(univariate_tbl_1)
choose1 <- if (univariate_tbl_1$dAIC[1] < -2) univariate_tbl_1$term[1] else NULL
predictors_1 <- if (is.null(choose1)) stop("STOP, no new predictor chosen") else setdiff(predictors_clean2, choose1)

```

```{r forward_selection_2}

base_rhs_1 <- as.formula(paste("~", choose1))
base_flist_1 <- build_flist_linear(base_rhs_1, K)
aic_base_1 <- AIC(mgcv::gam(base_flist_1, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude))

univariate_tbl_2 <- lapply(predictors_1, function(x){
  f1 <- update(base_rhs_1, as.formula(paste("~ . +", x)))
  list_f <- build_flist_linear(f1, K)
  m <- gam(list_f, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)
  data.frame(term = x, AIC  = AIC(m), dAIC = AIC(m) - aic_base_1)
}) %>% bind_rows() %>% arrange(AIC)
head(univariate_tbl_2)
choose2 <- if (univariate_tbl_2$dAIC[1] < -2) univariate_tbl_2$term[1] else NULL
predictors_2 <- if (is.null(choose2)) stop("STOP, no new predictor chosen") else setdiff(predictors_1, choose2)

```

```{r forward_selection_3}

base_rhs_2 <- update(base_rhs_1, as.formula(paste("~ . +", choose2)))
base_flist_2 <- build_flist_linear(base_rhs_2, K)
aic_base_2 <- AIC(mgcv::gam(base_flist_2, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude))

univariate_tbl_3 <- lapply(predictors_2, function(x){
  f1 <- update(base_rhs_2, as.formula(paste("~ . +", x)))
  list_f <- build_flist_linear(f1, K)
  m <- gam(list_f, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)
  data.frame(term = x, AIC  = AIC(m), dAIC = AIC(m) - aic_base_2)
}) %>% bind_rows() %>% arrange(AIC)
head(univariate_tbl_3)
choose3 <- if (univariate_tbl_3$dAIC[1] < -2) univariate_tbl_3$term[1] else NULL
predictors_3 <- if (is.null(choose3)) stop("STOP, no new predictor chosen") else setdiff(predictors_2, choose3)

```

```{r forward_selection_4}

base_rhs_3 <- update(base_rhs_2, as.formula(paste("~ . +", choose3)))
base_flist_3 <- build_flist_linear(base_rhs_3, K)
aic_base_3 <- AIC(mgcv::gam(base_flist_3, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude))

univariate_tbl_4 <- lapply(predictors_3, function(x){
  f1 <- update(base_rhs_3, as.formula(paste("~ . +", x)))
  list_f <- build_flist_linear(f1, K)
  m <- gam(list_f, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)
  data.frame(term = x, AIC  = AIC(m), dAIC = AIC(m) - aic_base_3)
}) %>% bind_rows() %>% arrange(AIC)
head(univariate_tbl_4)
choose4 <- if (univariate_tbl_4$dAIC[1] < -2) univariate_tbl_4$term[1] else NULL
predictors_4 <- if (is.null(choose4)) stop("STOP, no new predictor chosen") else setdiff(predictors_3, choose4)

```

```{r forward_selection_5}

base_rhs_4 <- update(base_rhs_3, as.formula(paste("~ . +", choose4)))
base_flist_4 <- build_flist_linear(base_rhs_4, K)
aic_base_4 <- AIC(mgcv::gam(base_flist_4, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude))

univariate_tbl_5 <- lapply(predictors_4, function(x){
  f1 <- update(base_rhs_4, as.formula(paste("~ . +", x)))
  list_f <- build_flist_linear(f1, K)
  m <- gam(list_f, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)
  data.frame(term = x, AIC  = AIC(m), dAIC = AIC(m) - aic_base_4)
}) %>% bind_rows() %>% arrange(AIC)
head(univariate_tbl_5)
choose5 <- if (univariate_tbl_5$dAIC[1] < -2) univariate_tbl_5$term[1] else NULL
predictors_5 <- if (is.null(choose5)) stop("STOP, no new predictor chosen") else setdiff(predictors_4, choose5)

```

```{r forward_selection_6}

base_rhs_5 <- update(base_rhs_4, as.formula(paste("~ . +", choose5)))
base_flist_5 <- build_flist_linear(base_rhs_5, K)
aic_base_5 <- AIC(mgcv::gam(base_flist_5, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude))

univariate_tbl_6 <- lapply(predictors_5, function(x){
  f1 <- update(base_rhs_5, as.formula(paste("~ . +", x)))
  list_f <- build_flist_linear(f1, K)
  m <- gam(list_f, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)
  data.frame(term = x, AIC  = AIC(m), dAIC = AIC(m) - aic_base_5)
}) %>% bind_rows() %>% arrange(AIC)
head(univariate_tbl_6)
choose6 <- if (univariate_tbl_6$dAIC[1] < -2) univariate_tbl_6$term[1] else NULL # "average_household_income"
predictors_6 <- if (is.null(choose6)) stop("STOP, no new predictor chosen") else setdiff(predictors_5, choose6)

```

```{r forward_selection_7}

base_rhs_6 <- update(base_rhs_5, as.formula(paste("~ . +", choose6)))
base_flist_6 <- build_flist_linear(base_rhs_6, K)
aic_base_6 <- AIC(mgcv::gam(base_flist_6, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude))

univariate_tbl_7 <- lapply(predictors_6, function(x){
  f1 <- update(base_rhs_6, as.formula(paste("~ . +", x)))
  list_f <- build_flist_linear(f1, K)
  m <- gam(list_f, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)
  data.frame(term = x, AIC  = AIC(m), dAIC = AIC(m) - aic_base_6)
}) %>% bind_rows() %>% arrange(AIC)
head(univariate_tbl_7)
choose7 <- if (univariate_tbl_7$dAIC[1] < -2) univariate_tbl_7$term[1] else NULL # "average_household_income"
predictors_7 <- if (is.null(choose7)) stop("STOP, no new predictor chosen") else setdiff(predictors_6, choose7)

```


## Selected model

```{r best_model}

best_rhs <- base_rhs_6
best_flist <- build_flist_linear(best_rhs, K)
best_model <- mgcv::gam(best_flist, data = df, family = mgcv::multinom(K = K), method = "REML", select = TRUE, na.action = na.exclude)
print(summary(best_model))

cat("\nBest_model AIC:\n")
print(AIC(best_model))

```

95%CIs from model summary SE:
```{r model_95pct_CI}

# extract parametric coefficient table
coef_summary <- summary(best_model) # model summary
coef_betas <- coef_summary$p.coeff # just betas
coef_SEs <- coef_summary$se # just standard errors
param_SEs <- coef_SEs[names(coef_SEs) %in% names(coef_betas)] # keep only SE values with names matching the parametric coefficient names (not spatial smooths)
coef_ps <- coef_summary$p.pv # just p-values
# 95% CI using normal approximation
zcrit <- qnorm(1 - 0.05/2)  # 1.96 for 95% CI
# make summary table with 95%CI
coef_tab_ci <- cbind(
  coef_betas,
  param_SEs,
  CI_lower = coef_betas - zcrit * param_SEs,
  CI_upper = coef_betas + zcrit * param_SEs,
  coef_ps
)
coef_tab_ci

```

#### Confusion martrix

Confusion matrix: observed vs predicted clusters
```{r confusion_matrix}

levs <- levels(wmel_model_df_scaled$cluster)                                                              # assign levels
predicted_probs <- predict(best_model, type = "response")                                                  # calculate residuals per cluster
colnames(predicted_probs) <- levs                                                                         # assign colnames
predicted_class <- factor(colnames(predicted_probs)[max.col(predicted_probs, ties.method = "first")], levels = levs)  # predicted clusters
observed_class <- factor(wmel_model_df_scaled$cluster, levels = levs)                                    # observed clusters
confusion_matrix <- caret::confusionMatrix(predicted_class, observed_class)
confusion_matrix

```

#### Baseline probabilities of each trajectory group

Average predictors and trajectories by release zone:
```{r baseline_probabilities_by_release_zone}

# scaled - must be scaled to provide interpretable outputs from predict() because the model was fitted to scaled data
scaled_baseline_data <- df %>%
  group_by(release_zone) %>%
  summarise(proportion_buildings_over6m = median(proportion_buildings_over6m, na.rm = TRUE),
            temp_c_mean_idw = median(temp_c_mean_idw, na.rm = TRUE),
            proportion_households_no_sewage_collection = median(proportion_households_no_sewage_collection, na.rm = TRUE),
            human_population = median(human_population, na.rm = TRUE),
            elevation = median(elevation, na.rm = TRUE),
            water_tank_density = median(water_tank_density, na.rm = TRUE),
            neighbouring_wmel_pctg_mean_avg = median(neighbouring_wmel_pctg_mean_avg, na.rm = TRUE),
            .groups = "drop")
# calculate baseline probabilities of each trajctory based on covariate averages
pred_baseline <- predict(
  best_model,
  newdata = scaled_baseline_data,
  type = "response"
)
pred_average_baseline <- scaled_baseline_data %>%
  bind_cols(as.data.frame(pred_baseline)) %>%
  rename(
    Unsuccessful = V1,
    Low_Success = V2,
    Moderate_Success = V3,
    High_Success = V4
  )
pred_average_baseline

```

Average trajectories for whole deployment area:
```{r baseline_probabilities_by_release_area}

# What about overall probabilities not by release zone?
# create a tibble with probabilities and weights (number of cells per zone)
pred_by_zone <- tibble(
  release_zone = c("RJ1", "RJ2", "RJ3"),
  Unsuccessful   = c(0.102, 0.200, 0.285),
  Low_Success    = c(0.151, 0.401, 0.584),
  Moderate_Success = c(0.618, 0.356, 0.091),
  High_Success   = c(0.129, 0.044, 0.040),
  n_cells        = c(48, 42, 177)  # weights
)
# compute weighted average for each trajectory
area_wide_probs <- pred_by_zone %>%
  summarise(
    Unsuccessful    = sum(Unsuccessful    * n_cells) / sum(n_cells),
    Low_Success     = sum(Low_Success     * n_cells) / sum(n_cells),
    Moderate_Success = sum(Moderate_Success * n_cells) / sum(n_cells),
    High_Success    = sum(High_Success    * n_cells) / sum(n_cells)
  )
area_wide_probs

```

#### Spatial autocorrelation of model residuals

```{r spatial_autocorrelation_residuals}

# model residuals spatial autocorrelation
set.seed(1)
for (k in levs) {                                                                                          # Moran's I test
  r_k <- as.numeric(wmel_model_df_scaled$cluster == k) - predicted_probs[,k]
  # keep only rows with finite residuals
  ok <- is.finite(r_k)
  if (sum(ok) < 3) {
    cat("\nMoran's I for class:", k, " — skipped (<3 obs after NA removal)\n")
    next
  }
  # subset weights to the same indices
  r_use  <- r_k[ok]
  lw_use <- spdep::subset.listw(lw, ok)
  # calculate Moran's I
  cat("\nMoran’s I for class:", k, "\n")
  print(moran.mc(r_use, lw_use, nsim = 999))
}

```


Conclusion:
The sensitivity analysis suggests that uncertainty in the classification of a small number of trajectories influences which secondary predictors are retained during model selection, but does not alter building height and temperature which consistently emerged as the strongest predictors across both analyses.


