---
title: "multinomial_gam_sensitivity_forward_selection_with_resampling"
author: "K Tiley"
date: "2026-07-17"
output: html_document
---

```{r setup, include=FALSE}

library(dplyr)            # pipelines
library(tidyr)            # dataframe manipulation
library(reshape2)         # dataframe manipulation
library(forcats)          # fct_recode()
library(nnet)             # multinomial logistic regression
library(mgcv)             # GAMs
library(caret)            # regression
library(ggplot2)          # plotting
library(sf)               # spatial data
library(spatialsample)    # spatial cross-validation
library(spdep)            # spatial autocorrelation
library(car)              # multicollinearity
library(purrr)            # iteration (map_* etc.)
library(rlang)
library(tibble)
library(stringr)
library(ggbeeswarm)
library(parallel)

```


## Prepare data

```{r data_prep}

# upload predictors
predictors_df <- read.csv("model_predictors_df_updatedflightbarriers2025-09-30.csv", header = TRUE)
#names(predictors_df)

# upload outcome trajectories
wmel_reassigned <- read.csv("k4_reassigned_wmel_all_2025-09-24.csv", header = TRUE)  # outcome with reassignment for decreasing slopes

# prepare model dataframes
wmel_clustered_num <- wmel_reassigned %>%
  filter(!is.na(cluster)) %>%                                                                               # remove rows where cluster is NA
  mutate(cluster = factor(cluster, levels = c("Unsuccessful","Low Success","Moderate Success","High Success")),
         cluster_num = dplyr::recode(as.character(cluster), "Unsuccessful" = 3L, "Low Success" = 4L, "Moderate Success" = 2L, "High Success" = 1L,
                                     .default = NA_integer_, .missing = NA_integer_),
         release_zone = factor(release_zone)) %>%
  distinct(cell.id, .keep_all = TRUE) %>%                                                                   # retain one row (and cluster label) per cell.id
  dplyr::select(cell.id, cluster, cluster_num, x, y, release_zone)

wmel_model_df <- wmel_clustered_num %>%
  dplyr::select(cell.id, cluster, release_zone) %>%
  left_join(predictors_df, by = "cell.id") %>%
  mutate(across(
    .cols = -c(cell.id, cluster, release_zone),
    .fns = ~ replace_na(., 0)
    ),
    release_zone5 = release_zone,
    release_zone = case_when(
      release_zone5 %in% c("RJ3.1", "RJ3.2", "RJ3.3") ~ "RJ3",
      release_zone5 == "RJ2" ~ "RJ2",
      TRUE ~ "RJ1"
    ),
    release_zone = factor(release_zone)
  )

# predictor pool (exclude id, response, coords)
predictors_og <- setdiff(
  names(wmel_model_df),
  c("cell.id", "cluster", "neighbouring_wmel_pctg_mean_avg", "release_zone", "release_zone5")
)

```


## Predictors

#### Tranformation for high Variance Inflation Factors (VIF)

As a rule of thumb:
-  <5 low, 5-10 moderate
-  >10 high multicollinearity & consider removing or combining
-  >30 severe multicollinearity & must address as model is not valid

Problem: proportion_residential & proportion_non_residential
Resolution: 'residualise' proportion_non_residential so that the non-residential buffer effect is retained without the non-residential data trends explained by proportion_residential
How: fit a model to get residuals, but which model? -> evidence of non-linearity, so fit a GAM.

```{r pred_vif_refine}

aux <- mgcv::gam(proportion_non_residential ~ s(proportion_residential, k = 6), data = wmel_model_df, method = "REML")
wmel_model_df$proportion_non_residential <- resid(aux)

```


## Scale predictors

Scale predictors for comparable coefficients which tell you the expected change in log-odds per 1 SD increase in the predictor. Method: subtract predictor mean and divide by SD.

```{r pred_scaled}

predictors_clean2 <- c("proportion_buildings_over6m", "temp_c_mean_idw", "elevation", "rain_24h_mean_idw")
wmel_model_df_scaled <- wmel_model_df[, c(predictors_clean2, "cell.id", "cluster", "release_zone", "neighbouring_wmel_pctg_mean_avg")] %>%
  mutate(across(all_of(predictors_clean2), ~ as.numeric(scale(.x))))
#summary(wmel_model_df_scaled)

```


## Model data and helper function

```{r screening_df}

# final model dataframe
wmel_model_df_scaled_subset <- wmel_model_df_scaled %>%
  dplyr::select(any_of(c(predictors_clean2, "cluster", "release_zone", "neighbouring_wmel_pctg_mean_avg")))
df <- droplevels(wmel_model_df_scaled_subset)

df$cluster <- relevel(df$cluster, ref = "Unsuccessful")
df$cluster_code <- as.numeric(df$cluster) - 1L                                                            # recode response to integers 0..K with desired baseline as 0
K <- max(df$cluster_code)
f_list0 <- list(cluster_code ~ s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone), ~ s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone), ~s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone))
m0 <- gam(f_list0, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)  # intercept-only reference model
aic_base <- AIC(m0)

```

```{r screening_helpers}

build_flist_linear <- function(
    rhs, K, resp = "cluster_code",
    k = 10, bs = "ts"
    ){

  # rhs  = RHS formula like ~ x1 + x2 + x3
  # K    = number of classes
  # resp = response variable name
  # k    = default basis size (scalar, default 10) OR a *named* numeric vec per var
  # bs   = basis for s(), default "ts" (thin-plate w/ shrinkage)
  
  stopifnot(inherits(rhs, "formula"), K >= 2)
  
  term_labels <- xfun::attr2(terms(rhs), "term.labels")
  
  pick_k <- function(v) {
    if (length(k) > 1 && !is.null(names(k)) && v %in% names(k)) k[[v]] else as.numeric(k)[1]
  }
  
  # always add the spatial proxy smooth
#  spatial_proxy <- "s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10)"
#  spatial_proxy <- "s(release_zone, bs='re')"
  spatial_proxy <- "s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone)"


  # assemble RHS (deduplicate, preserve first occurrence)
  rhs_terms <- unique(c(term_labels, spatial_proxy))
  rhs_new   <- as.formula(paste("~", paste(rhs_terms, collapse = " + ")))
  
  # mgcv multinomial formula list: first has response, rest are RHS
  c(list(update(rhs_new, paste(resp, "~ ."))),
    rep(list(rhs_new), K - 1))
}

```


## Selected model

```{r best_model}

best_rhs <- ~ proportion_buildings_over6m + temp_c_mean_idw + elevation + rain_24h_mean_idw
best_flist <- build_flist_linear(best_rhs, K)
best_model <- mgcv::gam(best_flist, data = df, family = mgcv::multinom(K = K), method = "REML", select = TRUE, na.action = na.exclude)
print(summary(best_model))

cat("\nBest_model AIC:\n")
print(AIC(best_model))

```

95%CIs from model summary SE:
```{r model_95pct_CI}

# extract parametric coefficient table
coef_summary <- summary(best_model) # model summary
coef_betas <- coef_summary$p.coeff # just betas
coef_SEs <- coef_summary$se # just standard errors
param_SEs <- coef_SEs[names(coef_SEs) %in% names(coef_betas)] # keep only SE values with names matching the parametric coefficient names (not spatial smooths)
coef_ps <- coef_summary$p.pv # just p-values
# 95% CI using normal approximation
zcrit <- qnorm(1 - 0.05/2)  # 1.96 for 95% CI
# make summary table with 95%CI
coef_tab_ci <- cbind(
  coef_betas,
  param_SEs,
  CI_lower = coef_betas - zcrit * param_SEs,
  CI_upper = coef_betas + zcrit * param_SEs,
  coef_ps
)
coef_tab_ci

```

#### Confusion martrix

Confusion matrix: observed vs predicted clusters
```{r confusion_matrix}

levs <- levels(wmel_model_df_scaled$cluster)                                                              # assign levels
predicted_probs <- predict(best_model, type = "response")                                                  # calculate residuals per cluster
colnames(predicted_probs) <- levs                                                                         # assign colnames
predicted_class <- factor(colnames(predicted_probs)[max.col(predicted_probs, ties.method = "first")], levels = levs)  # predicted clusters
observed_class <- factor(wmel_model_df_scaled$cluster, levels = levs)                                    # observed clusters
confusion_matrix <- caret::confusionMatrix(predicted_class, observed_class)
confusion_matrix

```


## Sensitivity analysis: bootstrap stability analysis of forward selection

Resample raster cells with replacement over ≥500 iterations, re-running forward selection on each resample, and report the selection frequency of each of the 23 candidate covariates.

Helper function to re-run forward selection:
Start with null model -> Fit every univariate model -> Choose best -> Fit every remaining variable -> Choose best -> Repeat until ΔAIC > -2

```{r forward_selection_helper}

forward_select <- function(data,
                           predictors,
                           K){

    selected <- character(0)
    remaining <- predictors
    current_formula <- ~1
    current_model <- gam(
        build_flist_linear(current_formula, K),
        data = data,
        family = mgcv::multinom(K = K),
        method = "ML",
        select = TRUE
    )

    current_AIC <- AIC(current_model)
    
    history <- data.frame(step = integer(),
                          variable = character(),
                          AIC = numeric(),
                          dAIC = numeric(),
                          stringsAsFactors = FALSE)
    
    repeat{
      if(length(remaining) == 0)
        break
      
      results <- lapply(remaining,function(v){
        
        test_formula <- update(current_formula, paste(". ~ . +", v))
        
        fit <- tryCatch({
          
          model <- gam(
            build_flist_linear(test_formula, K),
            data = data,
            family = mgcv::multinom(K = K),
            method = "ML",
            select = TRUE)
          
          data.frame(variable = v,
                     AIC = AIC(model))
          
          }, error = function(e){
            
            data.frame(variable = v,
                       AIC = Inf)
            
            })
        
        fit
        
        }) |>
        bind_rows()
      
      if(all(!is.finite(results$AIC)))
        break
        
        results$dAIC <- results$AIC - current_AIC
        best <- results[which.min(results$AIC), ]
        improvement <- current_AIC - best$AIC
        if(improvement < 2)
            break
        
        history <- rbind(history,
                         data.frame(step = nrow(history) + 1,
                                    variable = best$variable,
                                    AIC = best$AIC,
                                    dAIC = improvement))
        selected <- c(selected, best$variable)
        remaining <- setdiff(remaining, best$variable)
        current_formula <- update(current_formula, paste(". ~ . +", best$variable))
        current_AIC <- best$AIC
    }

    list(selected = selected,
         n_selected = length(selected),
         history = history,
         final_AIC = current_AIC)

}

```

```{r bootstrap_forward_selection_stability}

drops <- c("proportion_flood_prone","average_literacy","humidity_mean_idw","built_volume_density")
predictors_boot <- setdiff(predictors_og, drops)

set.seed(1234)
n_iterations <- 500
n_cores <- 8
cl <- makeCluster(n_cores)

clusterExport(cl, c("df", "predictors_boot", "K", "forward_select", "build_flist_linear"))

invisible(
clusterEvalQ(cl, {
  library(mgcv)
  library(dplyr)
  library(xfun)
})
)

boot_results <- parLapply(
  cl,
  seq_len(n_iterations),
  function(i){

    boot_df <- df[sample.int(nrow(df), replace = TRUE), ]

    tryCatch({

      res <- forward_select(data = boot_df,
                            predictors = predictors_boot,
                            K = K)
      res$iteration <- i
      res

    }, error = function(e){

      list(selected = character(0),
           n_selected = NA_integer_,
           history = NULL,
           final_AIC = NA_real_,
           iteration = i,
           error = conditionMessage(e))
      
    })
  }
)

stopCluster(cl)

```


```{r samples_converged}

# How many bootstrap samples converged?

boot_results <- Filter(function(x) is.null(x$error), boot_results)
n_success <- length(boot_results)
n_failed <- n_iterations - n_success
cat(n_success, "of", n_iterations, "bootstrap samples converged;", n_failed, "failed.\n")

```

```{r selection_frequency}

# In how many bootstrap samples was each covariate selected?

selected_vars <- lapply(boot_results, `[[`, "selected")

selection_counts <- tibble(variable = unlist(selected_vars)) |>
  count(variable, name = "times_selected")

selection_frequency <- tibble(variable = predictors_boot) |>
  left_join(selection_counts, by = "variable") |>
  mutate(times_selected = coalesce(times_selected, 0L),
         frequency = round(times_selected / n_success, 3)) |>
  arrange(desc(frequency), variable)

selection_frequency

```

```{r n_selected}

# How many bootstrap samples retained # number of variables?

n_selected_summary <- tibble(n_selected = sapply(boot_results, `[[`, "n_selected")) |>
  count(n_selected, name = "bootstrap_samples") |>
  mutate(percent = round(100 * bootstrap_samples / sum(bootstrap_samples), 1))

n_selected_summary

```

```{r boot_stability}

# How stable is the ordering?

first_selected <- tibble(first_variable = sapply(boot_results, function(x) x$history$variable[1])) |>
  count(first_variable, name = "times_first_selected") |>
  mutate(frequency = round(times_first_selected / n_success, 3)) |>
  arrange(desc(times_first_selected))

first_selected

```

```{r boot_stepwise}

# What does step by step selection look like?

step_summary <- step_history |>
  count(step, variable, name = "times_selected") |>
  group_by(step) |>
  mutate(frequency = round(times_selected / sum(times_selected), 3)) |>
  arrange(step, desc(times_selected))

step_summary

```


Conclusion:
Overall, the bootstrap sensitivity analysis indicates that the principal findings are not driven by instability of the forward-selection procedure. The strongest support was consistently observed for the proportion of tall buildings and average air temperature, while elevation and rainfall showed more moderate selection stability.


