---
title: "mulitnomial_gam_bootstrapped_model_selection"
author: "K Tiley"
date: "2026-05-25"
output: html_document
---

```{r setup, include=FALSE}

library(dplyr)            # pipelines
library(tidyr)            # dataframe manipulation
library(reshape2)         # dataframe manipulation
library(forcats)          # fct_recode()
library(nnet)             # multinomial logistic regression
library(mgcv)             # GAMs
library(caret)            # regression
library(ggplot2)          # plotting
library(sf)               # spatial data
library(spatialsample)    # spatial cross-validation
library(spdep)            # spatial autocorrelation
library(car)              # multicollinearity
library(purrr)            # iteration (map_* etc.)
library(rlang)
library(tibble)
library(stringr)
library(ggbeeswarm)

```


## Prepare data

```{r data_prep}

# upload predictors
predictors_df <- read.csv("model_predictors_df_updatedflightbarriers2025-09-30.csv", header = TRUE)
#names(predictors_df)

# upload outcome trajectories
wmel_reassigned <- read.csv("k4_reassigned_wmel_all_2025-09-24.csv", header = TRUE)  # outcome with reassignment for decreasing slopes

# prepare model dataframes
wmel_clustered_num <- wmel_reassigned %>%
  filter(!is.na(cluster)) %>%                                                                               # remove rows where cluster is NA
  mutate(cluster = factor(cluster, levels = c("Unsuccessful","Low Success","Moderate Success","High Success")),
         cluster_num = dplyr::recode(as.character(cluster), "Unsuccessful" = 3L, "Low Success" = 4L, "Moderate Success" = 2L, "High Success" = 1L,
                                     .default = NA_integer_, .missing = NA_integer_),
         release_zone = factor(release_zone)) %>%
  distinct(cell.id, .keep_all = TRUE) %>%                                                                   # retain one row (and cluster label) per cell.id
  dplyr::select(cell.id, cluster, cluster_num, x, y, release_zone)

wmel_model_df <- wmel_clustered_num %>%
  dplyr::select(cell.id, cluster, release_zone) %>%
  left_join(predictors_df, by = "cell.id") %>%
  mutate(across(
    .cols = -c(cell.id, cluster, release_zone),
    .fns = ~ replace_na(., 0)
    ),
    release_zone5 = release_zone,
    release_zone = case_when(
      release_zone5 %in% c("RJ3.1", "RJ3.2", "RJ3.3") ~ "RJ3",
      release_zone5 == "RJ2" ~ "RJ2",
      TRUE ~ "RJ1"
    ),
    release_zone = factor(release_zone)
  )

# predictor pool (exclude id, response, coords)
predictors_og <- setdiff(
  names(wmel_model_df),
  c("cell.id", "cluster", "neighbouring_wmel_pctg_mean_avg", "release_zone", "release_zone5")
)

```


## Predictors

#### Tranformation for high Variance Inflation Factors (VIF)

As a rule of thumb:
-  <5 low, 5-10 moderate
-  >10 high multicollinearity & consider removing or combining
-  >30 severe multicollinearity & must address as model is not valid

Problem: proportion_residential & proportion_non_residential
Resolution: 'residualise' proportion_non_residential so that the non-residential buffer effect is retained without the non-residential data trends explained by proportion_residential
How: fit a model to get residuals, but which model? -> evidence of non-linearity, so fit a GAM.

```{r pred_vif_refine}

aux <- mgcv::gam(proportion_non_residential ~ s(proportion_residential, k = 6), data = wmel_model_df, method = "REML")
wmel_model_df$proportion_non_residential <- resid(aux)

```


## Scale predictors

Scale predictors for comparable coefficients which tell you the expected change in log-odds per 1 SD increase in the predictor. Method: subtract predictor mean and divide by SD.

```{r pred_scaled}

predictors_clean2 <- c("proportion_buildings_over6m", "temp_c_mean_idw", "elevation", "rain_24h_mean_idw")
wmel_model_df_scaled <- wmel_model_df[, c(predictors_clean2, "cell.id", "cluster", "release_zone", "neighbouring_wmel_pctg_mean_avg")] %>%
  mutate(across(all_of(predictors_clean2), ~ as.numeric(scale(.x))))
#summary(wmel_model_df_scaled)

```


## Model data and helper function

```{r screening_df}

# final model dataframe
wmel_model_df_scaled_subset <- wmel_model_df_scaled %>%
  dplyr::select(any_of(c(predictors_clean2, "cluster", "release_zone", "neighbouring_wmel_pctg_mean_avg")))
df <- droplevels(wmel_model_df_scaled_subset)

df$cluster <- relevel(df$cluster, ref = "Unsuccessful")
df$cluster_code <- as.numeric(df$cluster) - 1L                                                            # recode response to integers 0..K with desired baseline as 0
K <- max(df$cluster_code)
f_list0 <- list(cluster_code ~ s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone), ~ s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone), ~s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone))
m0 <- gam(f_list0, data = df, family = mgcv::multinom(K = K), method = "ML", select = TRUE, na.action = na.exclude)  # intercept-only reference model
aic_base <- AIC(m0)

```

```{r screening_helpers}

build_flist_linear <- function(
    rhs, K, resp = "cluster_code",
    k = 10, bs = "ts"
    ){

  # rhs  = RHS formula like ~ x1 + x2 + x3
  # K    = number of classes
  # resp = response variable name
  # k    = default basis size (scalar, default 10) OR a *named* numeric vec per var
  # bs   = basis for s(), default "ts" (thin-plate w/ shrinkage)
  
  stopifnot(inherits(rhs, "formula"), K >= 2)
  
  term_labels <- xfun::attr2(terms(rhs), "term.labels")
  
  pick_k <- function(v) {
    if (length(k) > 1 && !is.null(names(k)) && v %in% names(k)) k[[v]] else as.numeric(k)[1]
  }
  
  # always add the spatial proxy smooth
#  spatial_proxy <- "s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10)"
#  spatial_proxy <- "s(release_zone, bs='re')"
  spatial_proxy <- "s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone)"


  # assemble RHS (deduplicate, preserve first occurrence)
  rhs_terms <- unique(c(term_labels, spatial_proxy))
  rhs_new   <- as.formula(paste("~", paste(rhs_terms, collapse = " + ")))
  
  # mgcv multinomial formula list: first has response, rest are RHS
  c(list(update(rhs_new, paste(resp, "~ ."))),
    rep(list(rhs_new), K - 1))
}

```

```{r forward_selection_helper}

run_forward_selection <- function(df_in, predictors_clean2, K, aic_base) {

  df <- droplevels(df_in)

  df$cluster <- relevel(df$cluster, ref = "Unsuccessful")
  df$cluster_code <- as.numeric(df$cluster) - 1L

  # base model
  f_list0 <- list(
    cluster_code ~ s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone),
    ~ s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone),
    ~ s(neighbouring_wmel_pctg_mean_avg, bs='ts', k=10, by=release_zone)
  )

  m0 <- gam(
    f_list0,
    data = df,
    family = mgcv::multinom(K = K),
    method = "ML",
    select = TRUE,
    na.action = na.exclude
  )

  selected <- c()

  remaining <- predictors_clean2

  repeat {

    best_term <- NULL
    best_dAIC <- 0

    for (x in remaining) {

      f1 <- as.formula(paste("~", paste(c(selected, x), collapse = " + ")))
      list_f <- build_flist_linear(f1, K)

      m <- gam(
        list_f,
        data = df,
        family = mgcv::multinom(K = K),
        method = "ML",
        select = TRUE,
        na.action = na.exclude
      )

      dAIC <- AIC(m) - aic_base

      if (dAIC < best_dAIC) {
        best_dAIC <- dAIC
        best_term <- x
      }
    }

    if (!is.null(best_term) && best_dAIC < -2) {
      selected <- c(selected, best_term)
      remaining <- setdiff(remaining, best_term)
    } else {
      break
    }
  }

  return(selected)
}

```

## Forward selection by AIC |>2|

```{r bootstrapped_forward_selection}

set.seed(123)

B <- 100
boot_selected <- vector("list", B)

n <- nrow(df)

start_time <- Sys.time()

for (b in 1:B) {
  
  if (b %% 5 == 0) cat("Bootstrap", b, "of", B, "\n")

  idx <- sample(1:n, size = n, replace = TRUE)
  df_b <- df[idx, ]

  boot_selected[[b]] <- tryCatch(
    run_forward_selection(
    df_in = df_b,
    predictors_clean2 = predictors_clean2,
    K = K,
    aic_base = aic_base
  ),

    error = function(e) {
      cat(
        "\nBootstrap", b,
        "failed:\n",
        conditionMessage(e), "\n"
      )

      return(NA)
    }
  )
}

end_time <- Sys.time()
end_time - start_time

```

```{r bootstrapped_forward_selection_outputs}

all_selected <- unlist(boot_selected)
names(all_selected) <- gsub("\\.1|\\.2", "", names(all_selected))
selection_freq <- sort(table(all_selected) / B, decreasing = TRUE)
selection_freq_df <- data.frame(
  covariate = names(selection_freq),
  freq = as.numeric(selection_freq)
)

head(selection_freq_df)

```

```{r case_deletion_bootstrapped_forward_selection}

set.seed(123)

B <- 100
boot_deletion_selected <- vector("list", B)

n <- nrow(df)

for (b in 1:B) {
  
  if (b %% 5 == 0) cat("Bootstrap", b, "of", B, "\n")

  idx <- sample(1:n, size = floor(0.9 * n), replace = TRUE)
  df_b <- df[idx, ]

  boot_deletion_selected[[b]] <- tryCatch(
    run_forward_selection(
    df_in = df_b,
    predictors_clean2 = predictors_clean2,
    K = K,
    aic_base = aic_base
  ),

    error = function(e) {
      cat(
        "\nBootstrap deletion", b,
        "failed:\n",
        conditionMessage(e), "\n"
      )

      return(NA)
    }
  )
}

```

```{r case_deletion_bootstrapped_forward_selection_outputs}

all_deletion_selected <- unlist(boot_deletion_selected)
names(all_deletion_selected) <- gsub("\\.1|\\.2", "", names(all_deletion_selected))
selection_deletion_freq <- sort(table(all_deletion_selected) / B, decreasing = TRUE)
selection_deletion_freq <- data.frame(
  covariate = names(selection_deletion_freq),
  freq = as.numeric(selection_deletion_freq)
)

head(selection_deletion_freq)

```

#### Spatially structured bootstrap

```{r spatial_bootstrapped_forward_selection}

set.seed(123)

zones <- unique(df$release_zone)
B <- 100
boot_spatial_selected <- vector("list", B)

n <- nrow(df)

for (b in 1:B) {
  
  if (b %% 5 == 0) cat("Bootstrap", b, "of", B, "\n")
  
  sampled_zones <- sample(zones, length(zones), replace = TRUE)
  df_b <- do.call(rbind, lapply(sampled_zones, function(z) {
    df[df$release_zone == z, ]
  }))

  boot_spatial_selected[[b]] <- tryCatch(
    run_forward_selection(
    df_in = df_b,
    predictors_clean2 = predictors_clean2,
    K = K,
    aic_base = aic_base
  ),

    error = function(e) {
      cat(
        "\nBootstrap spatial", b,
        "failed:\n",
        conditionMessage(e), "\n"
      )

      return(NA)
    }
  )
}

```

```{r spatial_bootstrapped_forward_selection_outputs}

all_spatial_selected <- unlist(boot_spatial_selected)
names(all_spatial_selected) <- gsub("\\.1|\\.2", "", names(all_spatial_selected))
selection_spatial_freq <- sort(table(all_spatial_selected) / B, decreasing = TRUE)
selection_spatial_freq <- data.frame(
  covariate = names(selection_spatial_freq),
  freq = as.numeric(selection_spatial_freq)
)

head(selection_spatial_freq)

```

