---
title: "wmel_pctg_k_clustering"
author: "K Tiley"
date: "2026-06-18"
output: html_document
---

Purpose of script is to load final predicted trajectories and compare k for k-means clustering.

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(cluster)
library(purrr)
library(tibble)
library(fpc)
library(patchwork)
library(ggalluvial)
```

## Load predicted trajectories

```{r load_data}

models_valid <- readRDS("wmel_pctg_binomial_glm_models_FittedRawTemp_PredictMeanTemp.rds")

prediction_matrix <- do.call(rbind, models_valid$predictions)
rownames(prediction_matrix) <- models_valid$cell.id
summary(as.vector(prediction_matrix))

```

## Fit k-means clustering for k = 3–6

```{r fit_k}

set.seed(42)
k_values <- 3:6

kmeans_results <- setNames(
  lapply(k_values, function(k) {
    kmeans(prediction_matrix, centers = k, nstart = 25)
  }),
  paste0("k", k_values)
)

models_by_k <- setNames(
  lapply(k_values, function(k) {
    out <- models_valid
    out$cluster <- kmeans_results[[paste0("k", k)]]$cluster
    out$k <- k
    out
  }),
  paste0("k", k_values)
)

```

## Sensitivity metrics: WSS, silhouette, and bootstrap stability

```{r k_metrics}

cluster_stability_summary <- purrr::map_dfr(k_values, function(k) {
  
  cb <- fpc::clusterboot(
    prediction_matrix,
    B = 500,
    bootmethod = "boot",
    clustermethod = fpc::kmeansCBI,
    k = k,
    seed = 42,
    count = FALSE
  )
  
  tibble(
    k = k,
    mean_jaccard_stability = mean(cb$bootmean, na.rm = TRUE),
    min_jaccard_stability = min(cb$bootmean, na.rm = TRUE)
  )
})

k_metrics <- tibble(
  k = k_values,
  total_withinss = purrr::map_dbl(kmeans_results, ~ .x$tot.withinss),
  avg_silhouette = purrr::map2_dbl(
    kmeans_results,
    k_values,
    ~ mean(cluster::silhouette(.x$cluster, dist(prediction_matrix))[, 3])
  )
) %>%
  left_join(cluster_stability_summary, by = "k")

k_metrics

```

```{r wss_plot, fig.height=5, fig.width=5, warning=FALSE}

wss_plot <- ggplot(k_metrics, aes(x = k, y = total_withinss)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_continuous(breaks = k_values) +
  labs(x = "", y = "Total WSS") +
  theme_minimal() +
  theme(axis.title.y = element_text(size = 14))

wss_plot

```

```{r sil_plot, fig.height=5, fig.width=5, warning=FALSE}

silhouette_plot <- ggplot(k_metrics, aes(x = k, y = avg_silhouette)) +
  geom_line() +
  geom_point(size = 2) +
  scale_x_continuous(breaks = k_values) +
  scale_y_continuous(breaks = c(0.4, 0.42, 0.44, 0.46, 0.48)) +
  labs(x = "Number of clusters (k)",
       y = "Average silhouette width") +
  theme_minimal() +
  theme(axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14))

silhouette_plot

```

```{r jaccard_plot, fig.height=6, fig.width=8, warning=FALSE}

cluster_stability_plot <- ggplot(cluster_stability_summary, aes(x = k)) +
  geom_line(aes(y = mean_jaccard_stability), linewidth = 1) +
  geom_point(aes(y = mean_jaccard_stability), size = 3) +
  geom_line(aes(y = min_jaccard_stability), linetype = "dashed", linewidth = 1) +
  geom_point(aes(y = min_jaccard_stability), size = 3) +
  geom_hline(yintercept = 0.75, linetype = "dotted") +
  geom_hline(yintercept = 0.60, linetype = "dotted") +
  annotate("text",
           x = min(cluster_stability_summary$k) - 0.05, y = 0.735,
           label = "Stable cluster threshold", size = 4.5) +
  annotate("text",
           x = min(cluster_stability_summary$k) - 0.05, y = 0.585,
           label = "Unstable cluster threshold", size = 4.5) +
  coord_cartesian(xlim = c(3, 6.8)) +
  labs(x = "", y = "Jaccard stability") +
#  labs(x = "Number of clusters (k)", y = "Jaccard stability",
#       title = "Bootstrapped stability of k-means solutions",
#       subtitle = "Solid = mean cluster stability; dashed = least stable cluster") +
  theme_minimal() +
  theme(axis.title.y = element_text(size = 14))

cluster_stability_plot

```

| Jaccard stability | Interpretation                            |
| ----------------- | ----------------------------------------- |
| > 0.85            | Highly stable cluster                     |
| 0.75–0.85         | Stable cluster                            |
| 0.60–0.75         | Moderately stable cluster                 |
| < 0.60            | Unstable cluster (often not reproducible) |

```{r supplementary_figure_panel, fig.height=5, fig.width=18, warning=FALSE}

supplementary_panel <- (wss_plot | silhouette_plot | cluster_stability_plot) +
  plot_layout(widths = c(5, 5, 8)) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 20, face = "bold"),
    plot.tag.position = c(0.02, 0.98)
  )

supplementary_panel
ggsave(filename = "FigureSupp_k-clusters_panel.jpeg", plot = supplementary_panel, width = 18, height = 5, dpi = 300)

```

## PCA visualisation of cluster separation

```{r pca_plot, fig.height=8, fig.width=10, warning=FALSE}

cluster_colours <- c("1" = "#eb5b2f", "2" = "#56d627", "3" = "#6ea3f5", "4" = "#e3a53b","5" = "deeppink", "6" = "mediumorchid4")

pca <- prcomp(prediction_matrix, scale. = TRUE)

pca_df <- tibble(
  cell.id = as.integer(rownames(prediction_matrix)),
  PC1 = pca$x[, 1],
  PC2 = pca$x[, 2]
)

pca_long <- purrr::map_dfr(seq_along(models_by_k), function(i) {
  model_k <- models_by_k[[i]]
  k_name <- names(models_by_k)[i]
  tibble(cell.id = model_k$cell.id,
         k = stringr::str_replace(k_name, "^k", "k = "),
         cluster = factor(model_k$cluster))
}) %>%
  left_join(pca_df, by = "cell.id")

pca_cluster_plot <- ggplot(pca_long, aes(x = PC1, y = PC2, colour = cluster)) +
  geom_point(size = 1.2, alpha = 0.8) +
  stat_ellipse(aes(group = cluster), level = 0.95, linewidth = 0.4) +
  facet_wrap(~ k) +
  scale_colour_manual(values = cluster_colours) +
  labs(x = "PC1", y = "PC2",
       colour = "Cluster",
       title = "PCA of predicted wMel trajectories by k-means cluster") +
  theme_minimal()

pca_cluster_plot

```

## Predicted trajectory plots for each k

```{r tradjectory_plot, fig.height=8, fig.width=10, warning=FALSE}

prediction_long <- as_tibble(prediction_matrix, rownames = "cell.id") %>%
  mutate(cell.id = as.integer(cell.id)) %>%
  pivot_longer(
    cols = -cell.id,
    names_to = "month_index",
    values_to = "wmel_pctg"
  ) %>%
  group_by(cell.id) %>%
  mutate(month_of_release = row_number()) %>%
  ungroup() %>%
  select(cell.id, month_of_release, wmel_pctg)

trajectory_long_by_k <- map_dfr(k_values, function(k) {
  prediction_long %>%
    left_join(
      models_by_k[[paste0("k", k)]] %>%
        select(cell.id, cluster),
      by = "cell.id"
    ) %>%
    mutate(
      k = paste0("k = ", k),
      cluster = factor(cluster)
    )
})

cluster_counts_by_k <- trajectory_long_by_k %>%
  distinct(k, cell.id, cluster) %>%
  count(k, cluster, name = "n_cells")

trajectory_plot_by_k <- ggplot(
  trajectory_long_by_k,
  aes(x = month_of_release, y = wmel_pctg, group = cell.id, colour = cluster)
) +
  geom_line(alpha = 0.4, linewidth = 0.4) +
  stat_summary(aes(group = cluster), fun = mean, geom = "line", colour = "black", linewidth = 0.6) +
  facet_grid(k ~ cluster, labeller = labeller(cluster = function(x) paste("Cluster", x))) +
  scale_colour_manual(values = cluster_colours) +
  labs(x = "Month of release", y = "Predicted proportion wMel",
       colour = "Cluster") +
  theme_minimal() +
  theme(legend.position = "none",
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        strip.text.x = element_text(size = 14),
        strip.text.y = element_text(size = 14))

trajectory_plot_by_k
#ggsave(filename = "FigureSupp_k-trajectory_panel.jpeg", plot = trajectory_plot_by_k, width = 10, height = 8, dpi = 300)

```

```{r tradjectory_maps, fig.height=8, fig.width=10, warning=FALSE}

spatial_base <- readRDS("wmel_with_temp.rds") %>%
  dplyr::select(cell.id, geometry) %>%
  st_as_sf()

cluster_df <- purrr::imap_dfr(models_by_k, function(df, name_k) { # flatten list
  df %>%
    dplyr::select(cell.id, cluster) %>%
    mutate(cell.id = as.integer(cell.id),
           k = name_k,
           cluster = factor(cluster))
})

map_data <- spatial_base %>%
  left_join(cluster_df, by = "cell.id") %>%
  filter(!is.na(k))

map_by_k_clusters <- ggplot(map_data) +
  geom_sf(aes(fill = cluster), color = NA) +
  facet_wrap(~k, ncol = 2) +
  scale_fill_manual(values = cluster_colours, na.value = "grey85") +
  theme_minimal() +
  labs(
    title = "Spatial distribution of k-means clusters across k values",
    fill = "Cluster"
  )

map_by_k_clusters

```


Save objects:

```{r save_objects, include=FALSE}

saveRDS(
  list(models_by_k = models_by_k,
       kmeans_results = kmeans_results,
       k_metrics = k_metrics,
       pca_long = pca_long,
       trajectory_long_by_k = trajectory_long_by_k),
  "wmel_kmeans_sensitivity_k3-6.rds"
)

```


## Interpretation

Silhouette: slightly favours k = 3.
WSS: favours increasing k.
Ecological trajectory interpretation: favours k = 4.
Bootstrap stability: k = 3 and k = 4 are both highly stable; k = 5 and 6 are not.

Conclusion:
Although k = 3 had the highest average silhouette score (0.47), the k = 4 solution was selected because it provided greater ecological resolution of trajectory types while maintaining high cluster stability (mean Jaccard = 0.84; minimum Jaccard = 0.81). In contrast, cluster stability declined substantially for k = 5 and k = 6, with at least one poorly reproducible cluster in each solution (minimum Jaccard = 0.24 and 0.54, respectively). These results indicate that k = 4 achieved a balance between ecological interpretability and statistical robustness.





## Reviewer request: Cell assignment changes across k

```{r alluvial_diagram_setup}

cluster_df <- bind_rows(
  models_by_k$k3,
  models_by_k$k4,
  models_by_k$k5,
  models_by_k$k6
) %>%
  dplyr::select(cell.id, k, cluster) %>%
  mutate(k = paste0("k", k))

cluster_wide <- cluster_df %>%
  pivot_wider(names_from = k,
              values_from = cluster)

```

```{r alluvial_diagram, fig.height=8, fig.width=12, warning=FALSE}

alluvial_df <- cluster_wide %>%
  group_by(k3, k4, k5, k6) %>%
  summarise(n = n(), .groups = "drop")

alluvial_plot <- ggplot(alluvial_df,
       aes(axis1 = k3, axis2 = k4, axis3 = k5, axis4 = k6, y = n)) +
  geom_alluvium(aes(fill = factor(k3)), alpha = 0.5) +
  geom_stratum(width = 1/5, fill = "white", color = "black") +
  geom_text(stat = "stratum",
            aes(label = paste0(after_stat(stratum), "\nn=", after_stat(count))),
            color = "black", size = 4) +
  scale_x_discrete(limits = c("k = 3", "k = 4", "k = 5", "k = 6"), expand = c(0.1, 0.1)) +
  labs(x = "Number of clusters (k)",
       y = "Number of raster cells (total n = 267)") +
  theme_bw() +
  theme(panel.grid = element_blank(),
        legend.position = "none",
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        strip.text.x = element_text(size = 14),
        strip.text.y = element_text(size = 14))

alluvial_plot
#ggsave(filename = "FigureSupp_k-alluvial_panel.jpeg", plot = alluvial_plot, width = 12, height = 8, dpi = 300)


```

```{r supplementary_figure_panel2, fig.height=16, fig.width=12, warning=FALSE}

supplementary_panel2 <- (alluvial_plot / trajectory_plot_by_k) +
  plot_layout(heights = c(8, 8)) +
  plot_annotation(tag_levels = "A") &
  theme(
    plot.tag = element_text(size = 20, face = "bold"),
    plot.tag.position = c(0.02, 0.98)
  )

supplementary_panel2
ggsave(filename = "FigureSupp_k-alluvial_trajectories_panel.jpeg", plot = supplementary_panel2, width = 12, height = 16, dpi = 300)

```

#### Cell assignment values

```{r cell_assignment}

cluster_assignments <- purrr::map_dfr(c(3, 4, 5), function(k) {
  models_by_k[[paste0("k", k)]] %>%
    dplyr::select(cell.id, cluster) %>%
    mutate(k = paste0("k", k))
}) %>%
  tidyr::pivot_wider(
    names_from = k,
    values_from = cluster
  )

```

```{r cell_assignment_34}

table_k3_k4 <- table(cluster_assignments$k3, cluster_assignments$k4)
prop_k3_k4_tbl <- prop.table(table_k3_k4, margin = 1) * 100
prop_k3_k4_tbl <- as.data.frame(prop_k3_k4_tbl) %>%
  dplyr::rename(
    k3_cluster = Var1,
    k4_cluster = Var2,
    percent = Freq
  ) %>%
  dplyr::mutate(percent = round(percent, 1)) %>%
  tidyr::pivot_wider(
    names_from = k4_cluster,
    values_from = percent,
    names_prefix = "k4_pctg_cluster_"
  ) %>%
  tibble::as_tibble()

cat("\n\n## For each k=3 cluster, where do its cells go under k=4? (% per k3 cluster)\n")
prop_k3_k4_tbl

cat("\n\n## Mean cell retention between k = 3 and k = 4\n")
mean_retention_k3_k4 <- prop_k3_k4_tbl %>%
  dplyr::select(-k3_cluster) %>%
  apply(1, max) %>%
  mean()
mean_retention_k3_k4

```

```{r cell_assignment_45}

table_k4_k5 <- table(cluster_assignments$k4, cluster_assignments$k5)
prop_k4_k5_tbl <- prop.table(table_k4_k5, margin = 1) * 100
prop_k4_k5_tbl <- as.data.frame(prop_k4_k5_tbl) %>%
  dplyr::rename(
    k4_cluster = Var1,
    k5_cluster = Var2,
    percent = Freq
  ) %>%
  dplyr::mutate(percent = round(percent, 1)) %>%
  tidyr::pivot_wider(
    names_from = k5_cluster,
    values_from = percent,
    names_prefix = "k5_pctg_cluster_"
  ) %>%
  tibble::as_tibble()

cat("\n\n## For each k=4 cluster, where do its cells go under k=5? (% per k4 cluster)\n")
prop_k4_k5_tbl

cat("\n\n## Mean cell retention between k = 4 and k = 5\n")
mean_retention_k4_k5 <- prop_k4_k5_tbl %>%
  dplyr::select(-k4_cluster) %>%
  apply(1, max) %>%
  mean()
mean_retention_k4_k5

```

Interpretation:
Going from k = 3 → k = 4, clusters 1 and 2 are largely preserved, but cluster 3 splits almost exactly 50:50 into two new groups.
Going from k = 4 → k = 5, clusters 1, 3 and 4 are highly stable (>90% retained in a single k=5 cluster), while cluster 2 is split into two substantial subgroups (44% and 54%).

Combined with the high bootstrapped stability of the k = 4 solution (mean Jaccard = 0.84), this supports k = 4 as a parsimonious balance between ecological resolution and cluster robustness.

