---
title: "wmel_pctg_trajectories"
author: "K Tiley"
date: "2026-06-19"
output: html_document
---

```{r setup, include=FALSE}

library(dplyr)
library(ggplot2)
library(tidyr)
library(purrr)
library(sf)

```

## Load k-means clustering outputs

```{r load_data}

km_out <- readRDS("wmel_kmeans_sensitivity_k3-6.rds")
models_by_k          <- km_out$models_by_k
kmeans_results       <- km_out$kmeans_results
k_metrics            <- km_out$k_metrics
trajectory_long_by_k <- km_out$trajectory_long_by_k

# just extract for 4 clusters as decided from k comparison metrics in wmel_pctg_k_clustering.Rmd
# explicitly ensure trajectory ordering

cluster_lookup <- c(
  "3" = "Unsuccessful",
  "1" = "Low Success",
  "2" = "Moderate Success",
  "4" = "High Success"
)

cluster_levels <- c(
  "Unsuccessful",
  "Low Success",
  "Moderate Success",
  "High Success"
)

models_valid4 <- models_by_k[["k4"]] %>%
  mutate(cluster = dplyr::recode(as.character(cluster), !!!cluster_lookup),
         cluster = factor(cluster, levels = cluster_levels))

trend_long <- trajectory_long_by_k  %>%
  filter(k == "k = 4") %>%
  mutate(cluster = dplyr::recode(as.character(cluster), !!!cluster_lookup),
         cluster = factor(cluster, levels = cluster_levels))

# sanity check number of cluster counts for both objects should be equal:
trend_counts <- trend_long %>%
  distinct(cell.id, cluster) %>%
  count(cluster)
model_counts <- models_valid4 %>%
  ungroup() %>%
  distinct(cell.id, cluster) %>%
  count(cluster)
# align and compare
all(trend_counts$n[match(trend_counts$cluster, model_counts$cluster)] == model_counts$n)

```


## Calculate trajectory slopes and reassign

```{r traj_slopes}

cell_slopes <- trend_long %>%
  group_by(cell.id) %>%
  summarise(
    slope = coef(lm(wmel_pctg ~ month_of_release))[2],
    .groups = "drop"
  )

```

Only reassign cells currently classified as Low, Moderate or High Success to Unsuccessful, based on slope <-0.005

```{r declining_slopes}

reassigned_ids <- cell_slopes %>%
  left_join(
    models_valid4 %>%
      select(cell.id, cluster),
    by = "cell.id"
  ) %>%
  filter(
    slope < -0.005,
    cluster %in% c("Low Success", "Moderate Success", "High Success")
  ) %>%
  pull(cell.id)

reassigned_ids

```

```{r reassign_declining_slopes}

trend_long_final <- trend_long %>%
  mutate(
    cluster = case_when(
      cell.id %in% reassigned_ids ~ "Unsuccessful",
      TRUE ~ as.character(cluster)
    ),
    cluster = factor(cluster, levels = levels(trend_long$cluster))
  )

models_valid4_final <- models_valid4 %>%
  left_join(cell_slopes, by = "cell.id") %>%
  mutate(
    cluster = case_when(
      cell.id %in% reassigned_ids ~ "Unsuccessful",
      TRUE ~ cluster
    )
  )

```

Sanity check: compare old vs new cluster assignments

```{r compare_old_vs_new}

old_counts <- models_valid4 %>%
  ungroup() %>%
  distinct(cell.id, cluster) %>%
  count(cluster, name = "old_n_cells")

new_counts <- models_valid4_final %>%
  ungroup() %>%
  distinct(cell.id, cluster) %>%
  count(cluster, name = "n_cells")

cluster_counts <- full_join(
  old_counts,
  new_counts,
  by = "cluster"
) %>%
  mutate(change = n_cells - old_n_cells,
         cluster = factor(cluster, levels = cluster_levels)) %>%
  arrange(cluster)
cluster_counts

```

## Plot trajectories

#### Figure 2. Trajectories in proportion wMel over release period

```{r traj_plot, fig.height=5, fig.width=9, warning=FALSE}

cluster_colours <- c(
  "Unsuccessful" = "#E6E482",
  "Low Success" = "#95C164",
  "Moderate Success" = "#428643",
  "High Success" = "#1B561F"
)

cluster_trajectories_plot <- ggplot(trend_long, aes(x = month_of_release, y = wmel_pctg, group = cell.id)) +
  geom_line(aes(colour = cluster), linewidth = 0.5, alpha = 0.4) +
  stat_summary(aes(group = cluster), fun = mean, geom = "line", colour = "black", linewidth = 0.9) +
  facet_wrap(~cluster, ncol = 4) +
  scale_color_manual(values = cluster_colours) +
  geom_text(data = cluster_counts, aes(label = paste0("n = ", old_n_cells)), x = 13, y = 1.04, inherit.aes = FALSE, size = 3.5) +
  coord_cartesian(ylim = c(0, 1.05)) +
  scale_x_continuous(breaks = seq(0, 25, by = 5)) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_text(size = 13)) +
  labs(x = "Month of Release", y = "Proportion wMel")

cluster_trajectories_plot

```

#### Supplementary Figure 1. Raster cell trajectories which were reassigned to unsuccessful
Reassigned cells highlighted:

```{r reassigned_traj_highlighted_plot, fig.height=5, fig.width=9, warning=FALSE}

trend_long_plot <- trend_long %>%
  mutate(plot_group = ifelse(cell.id %in% reassigned_ids, "Reassigned", as.character(cluster)))

cluster_trajectories_highlighted_plot <- ggplot(trend_long_plot, aes(x = month_of_release, y = wmel_pctg, group = cell.id)) +
  geom_line(aes(colour = plot_group), linewidth = 0.5, alpha = 0.4) +
  geom_line(data = subset(trend_long_plot, plot_group == "Reassigned"), colour = "red3", linewidth = 0.5, alpha = 1) + # reassigned cells (highlighted, solid + red)
  stat_summary(aes(group = cluster), fun = mean, geom = "line", colour = "black", linewidth = 0.9) +
  facet_wrap(~cluster, ncol = 4) +
  scale_color_manual(values = cluster_colours) +
  geom_text(data = cluster_counts, aes(label = paste0("n = ", old_n_cells)), x = 13, y = 1.04, inherit.aes = FALSE, size = 3.5) +
  coord_cartesian(ylim = c(0, 1.05)) +
  scale_x_continuous(breaks = seq(0, 25, by = 5)) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_text(size = 13)) +
  labs(x = "Month of Release", y = "Proportion wMel")

cluster_trajectories_highlighted_plot

```

```{r reassigned_cluster_map, fig.height=6, fig.width=8, warning=FALSE}

# cluster dataframe with coordinates
wmel_k4_reassigned <- read.csv("k4_reassigned_wmel_all_2025-09-24.csv")
map_plot_df <- wmel_k4_reassigned %>% filter(!is.na(cluster)) %>% group_by(cell.id) %>% slice(1) %>%
  mutate(cluster = factor(cluster, levels = c("Unsuccessful", "Low Success", "Moderate Success", "High Success")))
reassigned_cells <- map_plot_df %>%
  filter(cell.id %in% reassigned_ids)

# bbox for deployment area
bb_xy <- map_plot_df |>
  dplyr::distinct(cell.id, x, y) |>
  st_as_sf(coords = c("x","y"), crs = 32723) |>
  st_bbox()

# RJ city polygon
#muni_rj_ll <- geobr::read_municipality(code_muni = "RJ", year = 2020, simplified = TRUE)
#saveRDS(muni_rj_ll, "muni_rj_2020.rds")
muni_rj_ll <- readRDS("muni_rj_2020.rds")
rio_city_ll <- dplyr::filter(muni_rj_ll, name_muni == "Rio de Janeiro")
neighbour1_ll <- dplyr::filter(muni_rj_ll, name_muni == "São João de Meriti")
neighbour2_ll <- dplyr::filter(muni_rj_ll, name_muni == "Duque de Caxias")

# RJ state and other administrative polygons
#states_br_ll <- geobr::read_state(year = 2020, simplified = TRUE)
#saveRDS(states_br_ll, "states_br_2020.rds")
states_br_ll <- readRDS("states_br_2020.rds")
rj_state_ll  <- dplyr::filter(states_br_ll, abbrev_state == "RJ")
# fallback: rnaturalearth states (needs rnaturalearth + rnaturalearthdata)
#states_br_ll <- rnaturalearth::ne_states(country = "Brazil", returnclass = "sf")
#rj_state_ll  <- dplyr::filter(states_br_ll, name %in% c("Rio de Janeiro", "Rio de Janeiro State"))

# clusters map
clusters_map <- ggplot() +
  geom_sf(data = neighbour1_ll,  fill = "white", color = "grey40", linewidth = 0.2) +
  geom_sf(data = neighbour2_ll,  fill = "white", color = "grey40", linewidth = 0.2) +
  geom_sf(data = rio_city_ll,  fill = "white",     color = "black", linewidth = 0.5) +
  geom_raster(data = map_plot_df, aes(x = x, y = y, fill = cluster)) +
  geom_raster(data = reassigned_cells, aes(x = x, y = y), fill = "red3") +
  geom_text(aes(x = 686500, y = 7471500, label = "Atlantic\nOcean"), size = 3.8, fontface = "bold", color = "black") +
  geom_text(aes(x = 672000, y = 7470500, label = "Municipality of\nRio de Janeiro"), size = 3.8, fontface = "bold", color = "black") +
  coord_sf(crs = 32723,
           xlim = c(bb_xy["xmin"], bb_xy["xmax"] + 180),
           ylim = c(bb_xy["ymin"], bb_xy["ymax"] + 850)) +
  scale_fill_manual(values = cluster_colours, na.value = "grey70") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal() + 
  theme(
    panel.background = element_rect(fill = "#cfe8f3", colour = NA), # light blue ocean
    legend.position = c(0.98, 0.115),
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", colour = "grey60"),
    legend.box.margin = margin(4, 6, 4, 6),
    legend.key.height = unit(10, "pt"),
    legend.key.width  = unit(14, "pt")
  )
clusters_map

```


## Plot trajectories with reassigned cells

```{r reassigned_traj_plot, fig.height=5, fig.width=9, warning=FALSE}

cluster_trajectories_reassigned_plot <- ggplot(trend_long_final, aes(x = month_of_release, y = wmel_pctg, group = cell.id)) +
  geom_line(aes(colour = cluster), linewidth = 0.5, alpha = 0.4) +
  stat_summary(aes(group = cluster), fun = mean, geom = "line", colour = "black", linewidth = 0.9) +
  facet_wrap(~cluster, ncol = 4) +
  scale_color_manual(values = cluster_colours) +
  geom_text(data = cluster_counts, aes(label = paste0("n = ", n_cells)), x = 13, y = 1.04, inherit.aes = FALSE, size = 3.5) +
  coord_cartesian(ylim = c(0, 1.05)) +
  scale_x_continuous(breaks = seq(0, 25, by = 5)) +
  theme_minimal() +
  theme(legend.position = "none",
        strip.text = element_text(size = 13)) +
  labs(x = "Month of Release", y = "Proportion wMel")

cluster_trajectories_reassigned_plot

```

#### Manuscript Figure 3. Map of raster cells by replacement trajectory

```{r cluster_map, fig.height=6, fig.width=8, warning=FALSE}

# bbox for deployment area
bb_xy <- map_plot_df |>
  dplyr::distinct(cell.id, x, y) |>
  st_as_sf(coords = c("x","y"), crs = 32723) |>
  st_bbox()

# RJ city polygon
#muni_rj_ll <- geobr::read_municipality(code_muni = "RJ", year = 2020, simplified = TRUE)
#saveRDS(muni_rj_ll, "muni_rj_2020.rds")
muni_rj_ll <- readRDS("muni_rj_2020.rds")
rio_city_ll <- dplyr::filter(muni_rj_ll, name_muni == "Rio de Janeiro")
neighbour1_ll <- dplyr::filter(muni_rj_ll, name_muni == "São João de Meriti")
neighbour2_ll <- dplyr::filter(muni_rj_ll, name_muni == "Duque de Caxias")

# RJ state and other administrative polygons
#states_br_ll <- geobr::read_state(year = 2020, simplified = TRUE)
#saveRDS(states_br_ll, "states_br_2020.rds")
states_br_ll <- readRDS("states_br_2020.rds")
rj_state_ll  <- dplyr::filter(states_br_ll, abbrev_state == "RJ")
# fallback: rnaturalearth states (needs rnaturalearth + rnaturalearthdata)
#states_br_ll <- rnaturalearth::ne_states(country = "Brazil", returnclass = "sf")
#rj_state_ll  <- dplyr::filter(states_br_ll, name %in% c("Rio de Janeiro", "Rio de Janeiro State"))

# clusters map
clusters_map <- ggplot() +
  geom_sf(data = neighbour1_ll,  fill = "white", color = "grey40", linewidth = 0.2) +
  geom_sf(data = neighbour2_ll,  fill = "white", color = "grey40", linewidth = 0.2) +
  geom_sf(data = rio_city_ll,  fill = "white",     color = "black", linewidth = 0.5) +
  geom_raster(data = map_plot_df, aes(x = x, y = y, fill = cluster)) +
  geom_text(aes(x = 686500, y = 7471500, label = "Atlantic\nOcean"), size = 3.8, fontface = "bold", color = "black") +
  geom_text(aes(x = 672000, y = 7470500, label = "Municipality of\nRio de Janeiro"), size = 3.8, fontface = "bold", color = "black") +
  coord_sf(crs = 32723,
           xlim = c(bb_xy["xmin"], bb_xy["xmax"] + 180),
           ylim = c(bb_xy["ymin"], bb_xy["ymax"] + 850)) +
  scale_fill_manual(values = cluster_colours, na.value = "grey70") +
  labs(x = NULL, y = NULL, fill = NULL) +
  theme_minimal() + 
  theme(
    panel.background = element_rect(fill = "#cfe8f3", colour = NA), # light blue ocean
    legend.position = c(0.98, 0.115),
    legend.justification = c(1, 0),
    legend.background = element_rect(fill = "white", colour = "grey60"),
    legend.box.margin = margin(4, 6, 4, 6),
    legend.key.height = unit(10, "pt"),
    legend.key.width  = unit(14, "pt")
  )
clusters_map

```

